\Title

### 2022-09-27 Grimaldi-etal-BiolCybernetics.tex
Methods: 2.3

%%%-----------------------------------------------------------------
%: fig:motion_task
%%%-----------------------------------------------------------------
\begin{figure}[h!]%[ht!]
    \centering
    \includegraphics[width=0.95\linewidth]{figures/motion_task.pdf}
    \note{TODO: redo figure according to caption}
    \caption{
    {\bf Motion detection task.} {\bf (a)} To mimic the effect of a saccadic-like movement, the axis of view moves following a step-like random walk, and we show here on example path of the trajectory. {\bf (b)} We use large-scale natural images ($512\times512$) in which an aperture ($128\times128$) extracts a sub-image image in the axis of view (here the cross at the center of the image) such as to reproduce the effect of displacing the eye. {\bf (c)} Recording the dynamics of the sub-image as a function of time, it generates a naturalistic movie which may be transformed to an event-based representation. Mimicking the retina, this representation codes for proportional increments or decrements of the luminance in the image, respectively ON (in red) and OFF (in blue) events. This will constitute the input to the spiking neural network.}
    \label{fig:motion_task}
\end{figure}
%%%-----------------------------------------------------------------
Once these eye movement trajectories are generated, we can apply them to a visual scene. For this purpose, we selected a database of 10 natural grayscale images that are commonly used to study the statistics of natural images~\citep{olshausen_emergence_1996}. Note that these are pre-processed to equalize the energy in each frequency band (i.e., whitened). This process is known to occur as early as the retino-thalamic pathway~\citep{dan_efficient_1996}. These images are $512 \times 512$ in size and we will extract sub-images of size $128 \times 128$ which will be positioned around the center of the gaze at each time step (see \seeFig{motion_task}-(b)). We will discretize the time in $1~\ms$ bins and produce movies of duration N-T = $250~\ms$. To avoid border effects, we will position a draw of the complete trajectory at random in the image space so that the sub-image is translated using the position given by the trajectory at each time step. The translation is computed using a coordinate roll in the horizontal and vertical dimensions, followed by a sub-pixel translation defined in Fourier space~\citep{perrinet_sparse_2015}. Note that the magnitude of the displacement is relative to the time bin, and we have defined the velocity such that a velocity of $V=1$ corresponds to a movement of one pixel per frame (i.e., per time bin).

\note{ TODO: formula / say it is driven by image gradients / give scaling rules }
To transform each movie into events, we compute a gradient image (initialized at zero) by adding the gradient of the pixels' intensity over two successive frames. If, on a specific pixel at that specific timestamp, the absolute value of this gradient exceeds a threshold, an event is generated. The event has either an OFF or ON polarity, respectively whether the gradient is negative or positive. This signed threshold value is then subtracted from the residual gradient image. When applied to the whole movie, the event stream is as a consequence similar to the output of a neuromorphic camera~\citep{rasetto_challenges_2022}, that is, a list of events defined by $x_\arank$ and $y_\arank$ (their position on the pixel grid), their polarity $\polev_\arank$ (ON or OFF) and time $\timev_\arank$  (\seeFig{motion_task}-(c)). The goal here is to infer the correct motion solely by observing these events. 
This sensory signal representing the output of this event-based camera forms a discrete stream of events, $\event$ as defined above.   
Each event has an associated address, which is typically in the form $\presynaddr_\arank=(x_\arank, y_\arank, \polev_\arank)$. This defines a presynaptic address space $\presynaddrspace = [1, \Nx]\times[1, \Ny] \times [1, \Npol] \subset \mathbb{N}^3$ where $(\Nx, \Ny)$ is the size of the sensor in pixels and $\Npol$ is the number of polarities  ($\Npol=2$ for ON and OFF polarities). 
\note{ the task is full-field rigid translations, the model resolves it locally (analogy to the input from V1 to MT) }
%
\subsection{Extension as 3D convolutions for motion detection}
%%%-----------------------------------------------------------------
%
Having defined our task in the visual space for motion detection, we may extend the convolution-based detection model to a 3D convolution 
such that the representation would also benefit from spatial invariance. The use of spatiotemporal filters on a stream of events was shown to improve CNN performances for an action recognition task in~\cite{ghosh_spatiotemporal_2019} and we will here use a similar strategy, yet with a focus on using an event-based representation.
For that, we design 3D kernels of shape  $(\Kx, \Ky, \Ktime) = (9, 9, 11)$, respectively representing the two spatial dimensions and the range of delays.
Computations are performed on the spatio-temporal windows which are defined by the kernels and sliding around the events, that is, the center of the spatio-temporal window around the current event $\epsilon_\arank$. Note that to keep with a causal computation, the kernel are shifted in time, such that only past information is giving a response at the present time.
%
Finally, the output of the MLR model results in an event with the highest probability class, keeping the same timing as the event as input. Importantly, we include the prior knowledge that the event-based input is spatio-temporally clustered, and that these clusters are sparse. As a consequence, we only keep a given proportion of the most-active cells which we set here to $.01\%$ of all output voxels.
Given a spiking input, the model thus gives a spiking output  in the post-synaptic address space.

%
As the model is fully differentiable, we may now implement a supervised learning rule. The loss function of the MLR model is the binary cross entropy on the output of the classification layer and can be extended to the spatio-temporal convolutions. The supervision was implemented using the input binary events as defined above and the labels as the desired output. The labels were defined at each point in time as a one-hot encoding of the present motion in the channel corresponding to the current motion, and for all positions. Importantly, we weighted the loss functions only on the support corresponding to the active cells, such that the error was only back-propagated at the spatial locations of these active cells. Simulations are performed thanks to the PyTorch library using gradient descent with Adam (for $2^{12}$ movies and a learning rate of $10^{-5}$). We tested the effect of different parameters on a validation set to check the role of each parameter.

%
Finally, the output of the MLR model results in an event-based representation predicting at each position and time the probability for each motion. Such an output gives a form of optic flow which can be exploited for non-rigid motions, but here we defined for the sake of simplicity a readout method which applies to the full field. We have shown above that when different independent observations (here, motion estimated at different spatial locations) are known to have a common cause (here, the rigid motion of the image), then an optimal estimate of the logit of that probability is the sum of the logit of the independent probabilities. Taking the logit of the ouput probability given by the model, one can thus compute the probability for the output. This allows to compute the accuracy (as the percentage of times when the motion is exactly predicted) or the squared error $|| \speed - ||^2$ of the velocity estimate given by:
$
\speed = \sum_{\class} \speed_\class \cdot Pr(\class \vert image)
$ 
These computations are performed on a set of input which is different from that used at the training or validation stages. These metrics will be useful to compare our methods with other methods on our defined task.  The full code to reproduce the results of this paper are openly available at \url{https://github.com/SpikeAI/pyTERtorch}.  
%
\section{Results}
\label{sec:results}
%%%-----------------------------------------------------------------
%: fig:kernels
%%%-----------------------------------------------------------------
\subsection{Kernels learned for motion detection}
%%%-----------------------------------------------------------------
\begin{figure*}[ht!]
    {\centering
    %\vspace{-3cm}
    \includegraphics[width=\linewidth]{figures/motion_kernels.pdf}
    }
    %\vspace{-.5cm}
    \caption{
    	Representation of the weights for the $8$ learned kernels of the model corresponding to the OFF polarities and selective to the different motion directions
	(because of the symmetry observed between the ON and OFF event streams, kernels are similar for the ON polarities). These weights are associated to a specific delay on the \textit{delays} axis and to a presynaptic address defined on the two other axes.
	For the sake of clarity, the values in range $[-0.05, 0.05]$ are not shown. One sees positive (excitatory) coefficients for the specific direction of motion and negative (inhibitory) coefficients for all other directions.
	}
    \label{fig:kernels}
\end{figure*} 
%
%
After training our model, we first observe the weights learned for the different neurons (\seeFig{kernels}). Focusing on the positive weights, a strong selectivity is observed along specific axes for the different kernels. These directions can be easily associated to the direction of motion controlled in the motion clouds. For instance, the third and the seventh kernels show a horizontal selectivity to motion directions.
%
With the negative weights, one can observe an anti-selectivity for directions that do not correspond to the motion to which the kernel is selective to. This qualitative look at the 3D kernels allows the reader to infer for the $8$ different motion directions used to generate our synthetic event streams.

>>>>

%%%-----------------------------------------------------------------
%: fig:motion_task
%%%-----------------------------------------------------------------
\begin{figure}[h!]%[ht!]
    \centering
    \includegraphics[width=0.95\linewidth]{figures/motion_task.pdf}
    \note{TODO: redo figure according to caption}
    \caption{
    {\bf Motion detection task.} {\bf (a)} To mimic the effect of a saccadic-like movement, the axis of view moves following a step-like random walk, and we show here on example path of the trajectory. {\bf (b)} We use large-scale natural images ($512\times512$) in which an aperture ($128\times128$) extracts a sub-image image in the axis of view (here the cross at the center of the image) such as to reproduce the effect of displacing the eye. {\bf (c)} Recording the dynamics of the sub-image as a function of time, it generates a naturalistic movie which may be transformed to an event-based representation. Mimicking the retina, this representation codes for proportional increments or decrements of the luminance in the image, respectively ON (in red) and OFF (in blue) events. This will constitute the input to the spiking neural network.}
    \label{fig:motion_task}
\end{figure}
%%%-----------------------------------------------------------------
Once these eye movement trajectories are generated, we can apply them to a visual scene. For this purpose, we selected a database of 10 natural grayscale images that are commonly used to study the statistics of natural images~\citep{olshausen_emergence_1996}. Note that these are pre-processed to equalize the energy in each frequency band (i.e., whitened). This process is known to occur as early as the retino-thalamic pathway~\citep{dan_efficient_1996}. These images are $512 \times 512$ in size and we will extract sub-images of size $128 \times 128$ which will be positioned around the center of the gaze at each time step (see \seeFig{motion_task}-(b)). We will discretize the time in $1~\ms$ bins and produce movies of duration N-T = $250~\ms$. To avoid border effects, we will position a draw of the complete trajectory at random in the image space so that the sub-image is translated using the position given by the trajectory at each time step. The translation is computed using a coordinate roll in the horizontal and vertical dimensions, followed by a sub-pixel translation defined in Fourier space~\citep{perrinet_sparse_2015}. Note that the magnitude of the displacement is relative to the time bin, and we have defined the velocity such that a velocity of $V=1$ corresponds to a motion of one pixel per frame (i.e., per time bin).

\note{ TODO: formula / say it is driven by image gradients / give scaling rules }
To transform each movie into events, we compute a gradient image (initialized at zero) by adding the gradient of the pixels' intensity over two successive frames. If, on a specific pixel at that specific timestamp, the absolute value of this gradient exceeds a threshold, an event is generated. The event has either an OFF or ON polarity, respectively whether the gradient is negative or positive. This signed threshold value is then subtracted from the residual gradient image. When applied to the whole movie, the event stream is as a consequence similar to the output of a neuromorphic camera~\citep{rasetto_challenges_2022}, that is, a list of events defined by $x_\arank$ and $y_\arank$ (their position on the pixel grid), their polarity $\polev_\arank$ (ON or OFF) and time $\timev_\arank$  (\seeFig{motion_task}-(c)). The goal here is to infer the correct motion solely by observing these events. 
This sensory signal representing the output of this event-based camera forms a discrete stream of events, $\event$ as defined above.   
Each event has an associated address, which is typically in the form $\presynaddr_\arank=(x_\arank, y_\arank, \polev_\arank)$. This defines a presynaptic address space $\presynaddrspace = [1, \Nx]\times[1, \Ny] \times [1, \Npol] \subset \mathbb{N}^3$ where $(\Nx, \Ny)$ is the size of the sensor in pixels and $\Npol$ is the number of polarities  ($\Npol=2$ for ON and OFF polarities). 
\note{ the task is full-field rigid translations, the model resolves it locally (analogy to the input from V1 to MT) }
%
\subsection{Extension as 3D convolutions for motion detection}
%%%-----------------------------------------------------------------
%
Now that our motion detection task is defined in visual space, we can extend the convolution-based detection model to a 3D convolution so that the resulting model would also benefit from spatial invariance. The use of spatiotemporal filters on an event stream has improved the performance of CNNs for an action recognition task in~\cite{ghosh_spatiotemporal_2019} and here we will use a similar strategy, but focus on using an event-based representation.
For this purpose, we design 3D kernels of shape $(\Kx, \Ky, \Ktime) = (9, 9, 11)$, representing the two spatial dimensions and the range of time delays, respectively.
The computations are performed on the spatiotemporal windows that are defined by the kernels and are "moving" around the events, i.e. the spatial center of the spatiotemporal window around the current event $\epsilon_\arank$. Note that to remain within the framework of a causal calculation, the kernels are shifted in time such that only past information gives an answer at the present time.
%
The output of the MLR model corresponds to an event with the highest probability class, keeping the same timing as the input event. It is important to note that we include the prior knowledge that the event-based input is spatio-temporally clustered and that these clusters are sparse. Therefore, we will only select a given proportion of the most active cells, which we set here to $.01\%$ of all output voxels.
Given a stimulus input, the model thus yields a spike output in the postsynaptic address space.

%
Since the model is fully differentiable, we can now implement a supervised learning rule. The loss function of the MLR model is the binary cross-entropy on the output of the classification layer and can be extended to space-time convolutions. Supervision was implemented using the input binary events as defined above and the labels as the desired output. The labels were defined at each time point as a one-shot encoding of the current motion in the channel corresponding to the current motion for all positions. It is important to note that we weighted the cost function by considering only active cells, such that the error is only backpropagated to the spatial locations of these most active cells. Simulations are performed with the PyTorch library using gradient descent with Adam (for $2^{12}$ movies and a learning rate of $10^{-5}$). We tested the effect of different parameters on a validation set to quantify the role of each parameter.

%
Finally, the output of the MLR model is an event-based representation predicting at each position and at each time the probability of each motion. Such an output gives a form of optical flow that can be exploited for non-rigid motions, but we have defined here, for simplicity, an evaluation method that applies to our task with a full field motion. We have shown above that when different independent observations (here, the estimated motion at different spatial locations) are recognized as having a common cause (here, the rigid motion of the image), then an optimal estimate of the logit of this probability is the sum of the logits of the independent probabilities. By taking the logit of the probability of the output given by the model, we can therefore calculate the probability of the output. This allows one to calculate the accuracy (as the percentage of times the motion is accurately predicted) or the squared error $|| \speed - ||^2$ of the velocity estimate given by :
$
\speed = \sum_{\class} \speed_\class \cdot Pr(\class \vert image)
$ 
These calculations are performed on a different input dataset than the one used in the training or validation steps. These metrics will be useful to compare our methods with other methods on our defined task.  The complete code to reproduce the results of this paper is available at \url{https://github.com/SpikeAI/pyTERtorch}.  
%
\section{Results}
\label{sec:results}
%%%-----------------------------------------------------------------
%: fig:kernels
%%%-----------------------------------------------------------------
\subsection{Kernels learned for motion detection}
%%%-----------------------------------------------------------------
\begin{figure*}[ht!]
    {\centering
    %\vspace{-3cm}
    \includegraphics[width=\linewidth]{figures/motion_kernels.pdf}
    }
    %\vspace{-.5cm}
    \caption{
    	Representation of the weights for the $8$ learned kernels of the model corresponding to the OFF polarities and selective to the different motion directions
	(because of the symmetry observed between the ON and OFF event streams, kernels are similar for the ON polarities). These weights are associated to a specific delay on the \textit{delays} axis and to a presynaptic address defined on the two other axes.
	For the sake of clarity, the values in range $[-0.05, 0.05]$ are not shown. One sees positive (excitatory) coefficients for the specific direction of motion and negative (inhibitory) coefficients for all other directions.
	}
    \label{fig:kernels}
\end{figure*} 
%
%
After training our model, we first observe the weights learned for the different neurons (\seeFig{kernels}). Focusing on the positive weights, a strong selectivity is observed along specific axes for the different kernels. These directions can be easily associated to the direction of motion controlled in the motion clouds. For instance, the third and the seventh kernels show a horizontal selectivity to motion directions.
%
With the negative weights, one can observe an anti-selectivity for directions that do not correspond to the motion to which the kernel is selective to. This qualitative look at the 3D kernels allows the reader to infer for the $8$ different motion directions used to generate our synthetic event streams.
